Adaptive group-regularized logistic elastic net regression Supplementary material

### 2019-04-01 docs/supplement.Rnw
added all suggestions

  \section{Introduction}\label{sec:introduction}
  This document contains supplementary material to the paper `Adaptive 
  group-regularized logistic elastic net regression'. Section 
  \ref{sec:enetprior} contains extra details on the elastic net prior 
  distribution. Section \ref{sec:objectivefunction} gives the objective 
  function that is maximised in the variational Bayes procedure introduced in
  Section \ref{md-sec:estimation} of the Main Document (MD). Section
  \ref{sec:polyagamma} gives details on
  the P{\'o}lya-Gamma parametrisation introduced in Section 
  \ref{md-sec:estimation} of the MD. The
  derivations of the variational posterior used in the MD are shown in Section
  \ref{sec:variationaldistributions}. Section
  \ref{sec:computationalcomplexity} shows how we may reduce the computational
  complexity of our variational Bayes procedure. Section 
  \ref{sec:expectedloglikelihood} elaborates on the Variational Bayes expected
  joint log likelihood that is maximised to estimate the penalty multipliers as
  explained in Section \ref{md-sec:estimation} of the MD. The full procedure of
  the
  proposed method is shown in Section \ref{sec:fullestimationprocedure}. 
  Additional results to the applications in the MD Sections 
  \ref{md-sec:colorectal}--\ref{md-sec:oral}. We
  apply our method to a study on metabolomics in Alzheimer's in Section
  \ref{sec:alzheimer} and to study in microRNAs in cervical cancer data 
  in Section \ref{sec:cervical}. Finally, 
  we introduce the ridge and lasso versions of the proposed method in Section
  \ref{sec:evbridgelasso}.

@@

  Inspection of the updating equations in the MD, learns us that the 
  last term in the right-hand side of (\ref{eq:elboomega3}) equals zero, so that
  we are left with just (\ref{eq:elboomega4}). After a change of variables 
  $\psi_j = \tau_j - 1$, the two terms in (\ref{eq:elbobeta}) are as follows:
  \begin{subequations} \label{eq:elbobeta2}
	  \begin{align}
	    \E [\log \pi (\bbeta | \btau)] & = \frac{1}{2}\sum_{g=1}^G 
	    |\mathcal{G}(g)| \log \lambda'_g + \frac{p}{2} \log \lambda_2 - 
	    \frac{p}{2} \log (2 \pi) \\
	    & \,\,\,\,\, - \frac{\lambda_2}{2} \sum_{g=1}^G \lambda'_g \sum_{j \in 
	    \mathcal{G}(g)} \(1 + \frac{\lambda_1}{2 \sqrt{\lambda_2 \chi_j}}\) 
	    (\bmu_j^2 + \bSigma_{jj}) \\
	    & \,\,\,\,\, - \frac{1}{2} \sum_{j=1}^p \E [\log \psi_j] + \frac{1}{2} 
	    \sum_{j=1}^p \E [ \log(\psi_j + 1)], \\
	    \E [\log q (\bbeta )] & = -\frac{p}{2} - \frac{p}{2}\log (2 \pi) - 
	    \frac{p}{2} \log |\bSigma| .
	  \end{align}
  \end{subequations}
  We combine the terms in (\ref{eq:elbobeta2}) and arrive at the following for 
  (\ref{eq:elbobeta}):
  \begin{subequations} \label{eq:elbobeta3}
	   \begin{align}
	     \E_{Q^{(t)}} & [\log \pi (\bbeta | \btau)] - \E_{Q^{(t)}} 
	     [\log q (\bbeta )] \\ 
	     & = \frac{1}{2}\sum_{g=1}^G |\mathcal{G}(g)| \log \lambda'^{(t)}_g + 
	     \frac{p}{2} \log \lambda_2 + \frac{p}{2} \log |\bSigma^{(t)}| + 
	     \frac{p}{2} \\
	     & \,\,\,\,\, - \frac{\lambda_2}{2} \sum_{g=1}^G \lambda'^{(t)}_g 
	     \sum_{j \in \mathcal{G}(g)} \(1 + \frac{\lambda_1}{2 \sqrt{\lambda_2 
	     \chi_j^{(t)} } }\) \((\bmu^{(t)}_j)^2 + \bSigma_{jj}^{(t)}\) \\
	     & \,\,\,\,\, - \frac{1}{2} \sum_{j=1}^p \E_{Q^{(t)}} [\log \psi_j] + 
	     \frac{1}{2} \sum_{j=1}^p \E_{Q^{(t)}} [ \log(\psi_j + 1)].
	  \end{align}
  \end{subequations}

@@

  \section{Application to MicroRNAs in colorectal cancer}\label{sec:colorectal}
  \subsection{Additional comparisons}
  We compared \texttt{gren} to two methods in addition to the ones in the MD
  Section \ref{md-sec:colorectal}: 
  the sparse group lasso (\texttt{SGL}) by \cite{simon_sparse-group_2013} and 
  the group exponential lasso (\texttt{gel}) by \cite{breheny_group_2015}. For
  all methods with a tuning parameter $\alpha$, we set 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$ and present the AUC and Brier skill score 
  on the test set in Figures 
  \ref{fig:lines_micrornaseq_colorectal_cancer_res1_auc}--
  \ref{fig:lines_micrornaseq_colorectal_cancer_res1_briers}.
<<lines_micrornaseq_colorectal_cancer_res1_auc, fig.cap="Estimated AUC for the colorectal cancer example.", out.width="100%", fig.asp=1>>=
@

<<lines_micrornaseq_colorectal_cancer_res1_briers, fig.cap="Estimated Brier skill score for the colorectal cancer example.", out.width="100%", fig.asp=1>>=
@

@@

  \section{Application to RNAseq in oral cancer}
  \label{sec:oral}
  In addition to the results presented in Section \ref{md-sec:oral} of the 
  MD, we compared \texttt{gren} to the sparse group lasso (\texttt{SGL}) by 
  \cite{simon_sparse-group_2013} and the group exponential lasso (\texttt{gel}) 
  by \cite{breheny_group_2015}. The original group lasso methods are not 
  designed to deal with overlapping groups, so in addition to using the versions
  that allow for overlapping groups, we 
  cross-tabulated the two co-data partitions to create one paritioning with 25 
  groups for \texttt{cMCP} and \texttt{gel}.
  We use $\alpha \in \{0.05, 0.5, 0.95 \}$ for the
  methods that require this extra tuning parameter. The AUCs and Brier skill 
  scores calculated on the
  validation set are presented in Figures 
  \ref{fig:lines_rnaseq_lymph_node_metastasis_res1_auc}--
  \ref{fig:lines_rnaseq_lymph_node_metastasis_res1_briers}
<<lines_rnaseq_lymph_node_metastasis_res1_auc, fig.cap="Estimated AUC for the lymph node cancer example.", out.width="100%", fig.asp=1>>=
@

<<lines_rnaseq_lymph_node_metastasis_res1_briers, fig.cap="Estimated Briers skill score for the lymph node cancer example.", out.width="100%", fig.asp=1>>=
@

@@

  In this example, \texttt{GRridge} and \texttt{gren} underperform due to a 
  small number of features to learn the penalty parameters from. The other 
  methods estimate one global penalty parameter, so do not suffer from the 
  lower dimensionality of the problem.

@@

@@

  \section{Gibbs sampler}

  \begin{align*}
    \bbeta | \bomega, \bpsi & \sim \mathcal{N} (\bSigma, \bmu) \\
      & \text{with } \bSigma = [\X \tr \diag (\bomega_i) \X + 
      \lambda_2 \bLambda' + \lambda_2 \bLambda' \diag (\psi_j^{-1})]^{-1} 
      \text{ and } \bmu = \bSigma \X \tr (\y - \frac{1}{2} \mathbf{m}), \\
    \omega_i | \bbeta & \sim \mathcal{PG} (m_i, |\x_i \tr \bbeta|), \\
    \psi_j^{-1} | \beta_j & \sim \mathcal{IG} (\lambda_1/ (2 \lambda_2 
    \sqrt{\lambda'_{g(j)}} |\beta_j|), \lambda_1^2/(4 \lambda_2)).
  \end{align*}

>>>>

  \section{Introduction}\label{sec:introduction}
  This document contains supplementary material to the paper `Adaptive 
  group-regularized logistic elastic net regression'. Section 
  \ref{sec:enetprior} contains extra details on the elastic net prior 
  distribution. Section \ref{sec:objectivefunction} gives the objective 
  function that is maximised in the variational Bayes procedure introduced in
  Section \ref{md-sec:estimation} of the Main Document (MD). Section
  \ref{sec:polyagamma} gives details on
  the P{\'o}lya-Gamma parametrisation introduced in Section 
  \ref{md-sec:estimation} of the MD. The
  derivations of the variational posterior used in the MD are shown in Section
  \ref{sec:variationaldistributions}. Section
  \ref{sec:computationalcomplexity} shows how we may reduce the computational
  complexity of our variational Bayes procedure. Section 
  \ref{sec:expectedloglikelihood} elaborates on the Variational Bayes expected
  joint log likelihood that is maximised to estimate the penalty multipliers as
  explained in Section \ref{md-sec:estimation} of the MD. The full procedure of
  the
  proposed method is shown in Section \ref{sec:fullestimationprocedure}. 
  Extensions of our method are
  presented in Section \ref{sec:extensions}. The simulations in MD Section
  \ref{md-sec:simulations} are elaborated in Section \ref{sec:simulations}. 
  Additional results to the applications in the MD Sections 
  \ref{md-sec:colorectal}--\ref{md-sec:oral} are presented in Sections 
  \ref{sec:colorectal}--\ref{sec:oral}. We
  apply our method to a study on metabolomics in Alzheimer's in Section
  \ref{sec:alzheimer} and to study in microRNAs in cervical cancer data 
  in Section \ref{sec:cervical}. Finally, 
  we introduce the ridge and lasso versions of the proposed method in Section
  \ref{sec:evbridgelasso}.

@@

  Inspection of the updating equations (\ref{md-eq:VBupdateequations}) in the 
  MD, learns us that the 
  last term in the right-hand side of (\ref{eq:elboomega3}) equals zero, so that
  we are left with just (\ref{eq:elboomega4}). After a change of variables 
  $\psi_j = \tau_j - 1$, the two terms in (\ref{eq:elbobeta}) are as follows:
  \begin{subequations} \label{eq:elbobeta2}
	  \begin{align}
	    \E [\log \pi (\bbeta | \btau)] & = \frac{1}{2}\sum_{g=1}^G 
	    |\mathcal{G}(g)| \log \lambda'_g + \frac{p}{2} \log \lambda_2 - 
	    \frac{p}{2} \log (2 \pi) \\
	    & \,\,\,\,\, - \frac{\lambda_2}{2} \sum_{g=1}^G \lambda'_g \sum_{j \in 
	    \mathcal{G}(g)} \(1 + \frac{\lambda_1}{2 \sqrt{\lambda_2 \chi_j}}\) 
	    (\bmu_j^2 + \bSigma_{jj}) \\
	    & \,\,\,\,\, - \frac{1}{2} \sum_{j=1}^p \E [\log \psi_j] + \frac{1}{2} 
	    \sum_{j=1}^p \E [ \log(\psi_j + 1)], \\
	    \E [\log q (\bbeta )] & = -\frac{p}{2} - \frac{p}{2}\log (2 \pi) - 
	    \frac{p}{2} \log |\bSigma| .
	  \end{align}
  \end{subequations}
  We combine the terms in (\ref{eq:elbobeta2}) and arrive at the following for 
  (\ref{eq:elbobeta}):
  \begin{subequations} \label{eq:elbobeta3}
	   \begin{align}
	     \E_{Q^{(t)}} & [\log \pi (\bbeta | \btau)] - \E_{Q^{(t)}} 
	     [\log q (\bbeta )] \\ 
	     & = \frac{1}{2}\sum_{g=1}^G |\mathcal{G}(g)| \log \lambda'^{(t)}_g + 
	     \frac{p}{2} \log \lambda_2 + \frac{p}{2} \log |\bSigma^{(t)}| + 
	     \frac{p}{2} \\
	     & \,\,\,\,\, - \frac{\lambda_2}{2} \sum_{g=1}^G \lambda'^{(t)}_g 
	     \sum_{j \in \mathcal{G}(g)} \(1 + \frac{\lambda_1}{2 \sqrt{\lambda_2 
	     \chi_j^{(t)} } }\) \((\bmu^{(t)}_j)^2 + \bSigma_{jj}^{(t)}\) \\
	     & \,\,\,\,\, - \frac{1}{2} \sum_{j=1}^p \E_{Q^{(t)}} [\log \psi_j] + 
	     \frac{1}{2} \sum_{j=1}^p \E_{Q^{(t)}} [ \log(\psi_j + 1)].
	  \end{align}
  \end{subequations}

@@

	\subsection{Additional results}
	Additional results to MD Section \ref{md-sec:simulations} are given in 
	Figures \ref{fig:lines2_simulations_res1_auc_briers}-
	\ref{fig:lines2_simulations_res5_auc_briers}.

<<lines2_simulations_res1_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 1.", out.width="100%", fig.asp=1/2>>=
@

<<lines2_simulations_res2_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 2.", out.width="100%", fig.asp=1/2>>=
@

<<lines2_simulations_res3_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 3.", out.width="100%", fig.asp=1/2>>=
@

<<lines2_simulations_res4_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 4.", out.width="100%", fig.asp=1/2>>=
@

<<lines2_simulations_res5_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 5.", out.width="100%", fig.asp=1/2>>=
@

  \section{Application to MicroRNAs in colorectal cancer}\label{sec:colorectal}
  \subsection{Additional comparisons}
  We compared \texttt{gren} to two methods in addition to the ones in the MD
  Section \ref{md-sec:colorectal}: 
  the sparse group lasso (\texttt{SGL}) by \cite{simon_sparse-group_2013} and 
  the group exponential lasso (\texttt{gel}) by \cite{breheny_group_2015}. For
  all methods with a tuning parameter $\alpha$, we set 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$ and present the AUC and Brier skill score 
  on the test set in Figures 
  \ref{fig:lines_micrornaseq_colorectal_cancer_res1_auc}--
  \ref{fig:lines_micrornaseq_colorectal_cancer_res1_briers}.
<<lines_micrornaseq_colorectal_cancer_res1_auc, fig.cap="Estimated AUC for the colorectal cancer example with (a) $\\alpha=0.05$, (b) $\\alpha=0.5$, and (c) $\\alpha=0.95$.", fig.asp=1>>=
@

<<lines_micrornaseq_colorectal_cancer_res1_briers, fig.cap="Estimated Brier skill score for the colorectal cancer example with (a) $\\alpha=0.05$, (b) $\\alpha=0.5$, and (c) $\\alpha=0.95$.", out.width="100%", fig.asp=1>>=
@

@@

  \section{Application to RNAseq in oral cancer}
  \label{sec:oral}
  In addition to the results presented in Section \ref{md-sec:oral} of the 
  MD, we compared \texttt{gren} to the sparse group lasso (\texttt{SGL}) by 
  \cite{simon_sparse-group_2013} and the group exponential lasso (\texttt{gel}) 
  by \cite{breheny_group_2015}. The original group lasso methods are not 
  designed to deal with overlapping groups, so in addition to using the versions
  that allow for overlapping groups, we 
  cross-tabulated the two co-data partitions to create one paritioning with 25 
  groups for \texttt{cMCP} and \texttt{gel}.
  We use $\alpha \in \{0.05, 0.5, 0.95 \}$ for the
  methods that require this extra tuning parameter. The AUCs and Brier skill 
  scores calculated on the
  validation set are presented in Figures 
  \ref{fig:lines_rnaseq_oral_cancer_res1_auc}--
  \ref{fig:lines_rnaseq_oral_cancer_res1_briers}
<<lines_rnaseq_oral_cancer_res1_auc, fig.cap="Estimated AUC for the oral cancer example with (a) $\\alpha=0.05$, (b) $\\alpha=0.5$, and (c) $\\alpha=0.95$.", out.width="100%", fig.asp=1>>=
@

<<lines_rnaseq_oral_cancer_res1_briers, fig.cap="Estimated Briers skill score for the oral cancer example with (a) $\\alpha=0.05$, (b) $\\alpha=0.5$, and (c) $\\alpha=0.95$.", out.width="100%", fig.asp=1>>=
@

@@

  In this example, \texttt{GRridge} and \texttt{gren} underperform due to:
  (i) a small number of features to learn the penalty parameters from and (ii)
  Large negative correlations between the metabolites (see Figure 
  \ref{fig:heatmap_metabolomics_alzheimer_res1_cor}). Both (i) and (ii) make 
  estimation of the penalty multipliers difficult. The other 
  methods estimate one global penalty parameter, so do not suffer from the 
  lower dimensionality of the problem and the large negative correlation.

<<heatmap_metabolomics_alzheimer_res1_cor, fig.cap="Correlations between the metabolites.", out.width="50%", fig.asp=1>>=
library(Biobase)
library(lattice)
load("data/ESetMbolCSFPR2.Rdata")
pheno <- pData(ESetMbolCSFPR2)
metabol <- t(exprs(ESetMbolCSFPR2))
metabol.apoe <- metabol[(pheno$D_diag_name=="Probable AD" & 
                           pheno$APOE=="E4YES") |
                          (pheno$D_diag_name=="Subjectieve klachten" & 
                             pheno$APOE=="E4NO"), ]
cor.mat <- cor(scale(metabol.apoe))
levelplot(cor.mat, xlab=NULL, ylab=NULL, scales=list(draw=FALSE))
@

@@

  The estimated multipliers, AUC, and Brier skill score
  are in Figures \ref{fig:barplot_micrornaseq_cervical_cancer_res1}--
  \ref{fig:lines_micrornaseq_cervical_cancer_res1_briers}, respectively. In this
  example the group regularized methods outperform the regular methods with
  respect to discrimination (AUC) and calibration (Brier skill score).

@@

  % \section{Gibbs sampler}
  %   
  % \begin{align*}
  %   \bbeta | \bomega, \bpsi & \sim \mathcal{N} (\bSigma, \bmu) \\
  %     & \text{with } \bSigma = [\X \tr \diag (\bomega_i) \X + 
  %     \lambda_2 \bLambda' + \lambda_2 \bLambda' \diag (\psi_j^{-1})]^{-1} 
  %     \text{ and } \bmu = \bSigma \X \tr (\y - \frac{1}{2} \mathbf{m}), \\
  %   \omega_i | \bbeta & \sim \mathcal{PG} (m_i, |\x_i \tr \bbeta|), \\
  %   \psi_j^{-1} | \beta_j & \sim \mathcal{IG} (\lambda_1/ (2 \lambda_2 
  %   \sqrt{\lambda'_{g(j)}} |\beta_j|), \lambda_1^2/(4 \lambda_2)).
  % \end{align*}
